---
title: "Prior Predictive Checks with rethinking"
author: "Julian Frattini"
date: "2025-09-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(rethinking)
library(tidyverse)
library(patchwork)
```

This notebook demonstrates prior predictive checks with the `rethinking` package.

## Data

For demonstration, we can use the `rethinking::data("Howell1")` data, which contains "demographic data from the Kalahari !Kung San people collected by Nancy Howell" (see `vignette(Howell1)`).
For this simple problem, we focus on the subset of data points from adults (`age >= 18`).

```{r load-data}
data("Howell1")
d <- Howell1 %>% 
  filter(age >= 18)
```

For a better understanding of the data, we can visualize it in a simple scatterplot.
The relationship between `weight` (measured in kg) and `height` (measured in cm) is clearly positive.

```{r visualize-data}
ggplot(d, aes(x = weight, y = height)) +
  geom_point() +
  geom_smooth(method = "lm", formula = (y ~ x), se = FALSE, color = "grey") +
  labs(x = "Weight in kg", y = "Height in cm") +
  theme_bw()
```

## Prior Predictive Check

To model the height in relation to the weight, we have two options:

1. $height \sim \alpha + \beta*weight$, i.e., height is modeled by weight
2. $height \sim \alpha + \beta*(weight - \overline{weight})$, i.e., height is modeled by *weight deviation from the mean* $\overline{weight}$

The second option is preferable because here $\alpha$ represents the average height of a person with average weight.
In the first option, $\alpha$ would be the height of a person with weight = 0kg.
It is easier to pick priors for the former.

```{r calculate-average}
avg.weight = mean(d$weight)
```

### Choosing Priors

Richard McElreath proposes two different sets of priors[^1] which differ in their parameters for $\beta$.
In `priors1`, he selects a wide $\mathcal{N}(0, 10)$, in `priors2` only $log(\mathcal{N}(0, 1))$.

```{r select-priors}
# number of random draws
n = 100

priors1 <- data.frame(
  a = rnorm(n, 178, 20),
  b = rnorm(n, 0, 10)
) 

priors2 <- data.frame(
  a = rnorm(n, 178, 20),
  b = rlnorm(n, 0, 1)
) 
```

These priors mean that, on average, we assume that changing the `weight` variable by one unit changes the `height` variable by $\mathcal{N}(0, 10)$ or $log(\mathcal{N}(0, 1))$ units respectively.

### Visualizing Priors

To understand, whether these priors are feasible, we can visualize them.
For this, we define a function which visualizes the randomly sampled values for $\alpha$ and $\beta$ as lines in a plot.

```{r visualize-priors}
visualize.priors <- function(p) {
  ggplot(data = p) +
    geom_abline(aes(intercept = a-b*avg.weight, slope = b), color = "grey") +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_hline(yintercept = 272, linetype = "dashed") +
    xlim(30, 60) +
    ylim(0, 400) +
    labs(x = "Weight in kg", y = "Height in cm") +
    theme_bw()
}
```

With this visualization, we can compare the two priors.

```{r prior-predictive-check}
priorpc1 <- visualize.priors(priors1)
priorpc2 <- visualize.priors(priors2)

priorpc1 | priorpc2
```

The `priors1` clearly show unrealistic assumptions.
Some extreme slopes suggest that a person with 30kg weight might have a height of 400cm or more.
Other slopes suggest that, given a certain weight, the height of a person may be negative.
The `priors2` show a much more realistic scenario.
All slopes are now positive (which is achieved by taking a `lnorm` instead of a `norm` distribution) and much less steep (achieved by a variance of 1 instead of 10).
There are still a few extreme values, but the majority encode realistic assumptions.

[^1]: McElreath, R. (2018). Statistical rethinking: A Bayesian course with examples in R and Stan. Chapman and Hall/CRC.
