---
title: "Exercises: D-Separation"
author: "Julian Frattini"
date: "2025-04-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(patchwork)
# visualization and analysis of DAGs
library(ggdag)
```

This notebook contains some examples of applying d-separation and the backdoor criterion to derive an unbiased statistical model from a causal model.

### Example 3

This DAG is a variant of the M-bias with one additional edge $Z \rightarrow X$.

```{r example-3-dag}
dag3 <- dagify(
  a ~ c + d,
  b ~ a + e,
  d ~ c + e,
  exposure = "a", outcome = "b",
  coords = list(x = c(a = 0, b = 1, c = 0, d = 0.5, e = 1), 
                y = c(a = 0, b = 0, c = 1, d = 0.5, e = 1))
)
ggdag_status(dag3) + 
  guides(fill = "none", color = "none") +
  theme_dag()
```

The DAG contains two backdoor paths:

1. $a \leftarrow c \rightarrow d \leftarrow e \rightarrow b$
2. $a \leftarrow d \leftarrow e \rightarrow b$

The first backdoor path is closed by default because it contains a collider d.

```{r example-3-adjustment}
ggdag_adjustment_set(dag3) +
  theme(legend.position="bottom")
```

The adjustment set suggests two ways of de-confounding the effect of a on b. 
One way is to condition on e, which blocks the second backdoor path.
Another way is to condition on d.
However, this opens up the first backdoor path since it conditions on the collider d.
Additionally, one must condition on another variable on that path, e.g., c (or e, but then d would not be necessary in the first place).

### Example 4

```{r example-4-dag}
dag4 <- dagify(
  a ~ c,
  b ~ a + c + e,
  d ~ c + e,
  exposure = "a", outcome = "b",
  latent = c("c"),
  coords = list(x = c(a = 0, b = 1, c = 0.1, d = 0.5, e = 0.9), 
                y = c(a = 0, b = 0, c = -0.9, d = -1, e = -0.9))
)
ggdag_status(dag4) + 
  guides(fill = "none", color = "none") +
  theme_dag()
```

```{r example-4-adjustment}
ggdag_adjustment_set(dag4) +
  theme(legend.position="bottom")
```

```{r}
n <- 1000

d <- data.frame(
  c = rnorm(n, 0, 1),
  e = rnorm(n, 0, 1)
) %>% 
  mutate(
    d = rnorm(n, 2*c + 2*e, 1),
    a = rnorm(n, 2*c, 1),
    b = rnorm(n, 3*a + 2*c + 2*e)
  )
```

```{r}
lm(b ~ a, data = d)
lm(b ~ a + c, data = d)
lm(b ~ a + d, data = d)
lm(b ~ a + d + e, data = d)
```


