---
title: "Non-linear Effects"
author: "Julian Frattini"
date: "2026-02-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(patchwork)

library(brms)
library(marginaleffects)
```

This notebook demonstrates the recovery of non-linear effects.

## Data Simulation

For this demonstration, we simulate a simple data set with two causes `x` and `z`, and one effect `y`.
Notably, the effect of $x \ rightarrow y$ is not linear, but quadratic:

$$y \sim \mathcal{N}(\mu = 5 + 3x^2 -2z, \sigma = 0.5)$$

```{r simulate-data}
n <- 1e3
eff.x <- 3
eff.z <- -2

d <- data.frame(
  x = rnorm(n, mean = 0, sd = 1),
  z = rnorm(n, mean = 0, sd = 1)
) %>% 
  mutate(
    y = rnorm(n, 5 + eff.x*x^2 + eff.z*z, 0.5)
  )
```

For visual confirmation, we can plot the individual relationship between the two causes and the effect.

```{r visualize-data}
plt.xy <- ggplot(data = d, mapping = aes(x = x, y = y)) +
  geom_point() +
  theme_bw()

plt.zy <- ggplot(data = d, mapping = aes(x = x, y = z)) +
  geom_point() +
  theme_bw()

plt.xy | plt.zy
```

## Data Analysis

In the data analysis, we compare two models: a naive baseline assuming only linear relationships, and a more complex model assuming a quadratic relationship.
Quadratic predictors are encoded as `I(x^2)` in `brms`.

```{r formulae}
f1 <- bf(y ~ x + z)
f2 <- bf(y ~ I(x^2) + z)

get_prior(data = d, formula = f2, family = gaussian)
```

For the two models, we select uninformative priors.

```{r priors}
priors1 <- c(
  prior(normal(5, 2), class = Intercept),
  prior(normal(0, 2), class = b),
  prior(exponential(1), class = sigma)
)

priors2 <- c(
  prior(normal(5, 2), class = Intercept),
  prior(normal(0, 2), class = b),
  prior(normal(0, 1), class = b, coef = "IxE2"),
  prior(exponential(1), class = sigma)
)
```

Next, we perform a prior predictive check for the non-linear model to assure that the priors are feasible.

```{r prior-predictive-check}
p.nonlinear <- 
  brm(
    f2, data = d, family = gaussian, prior = priors2,
    iter = 2000, warmup = 1000, chains = 4, cores = 4,
    sample_prior="only",
    file = "../../fits/p.nonlinear", file_refit = "on_change"
  )

pp_check(p.nonlinear, ndraws = 100)
```

Next, we train both the linear and the non-linear model and perform a posterior predictive check to assert that the models have trained well.

```{r model-linear}
m.linear <- 
  brm(
    f1, data = d, family = gaussian, prior = priors1,
    iter = 2000, warmup = 1000, chains = 4, cores = 4,
    file = "../../fits/m.linear", file_refit = "on_change"
  )

pp_check(m.linear, ndraws = 100)
```

```{r model-nonlinear}
m.nonlinear <- 
  brm(
    f2, data = d, family = gaussian, prior = priors2,
    iter = 2000, warmup = 1000, chains = 4, cores = 4,
    file = "../../fits/m.nonlinear", file_refit = "on_change"
  )

pp_check(m.nonlinear, ndraws = 100)
```

## Evaluation

Finally, we can evaluate the two models.
First, we are going to look at the raw posterior distribution.
The plot contains a dashed line at the true effect.
It becomes very clear that the non-linear model approximates this true effect much better.

```{r comparison-posteriors}
post.x.linear <- summary(m.linear)$fixed["x",]
post.x.nonlinear <- summary(m.nonlinear)$fixed["IxE2",]

post.x <- rbind(post.x.linear, post.x.nonlinear)
cbind(factor = rownames(post.x), post.x) %>% 
  mutate(model = ifelse(factor == "x", "linear (x)", "quadratic (x^2)")) %>% 
  ggplot(aes(x = Estimate, y = model, color = factor(model))) +
    geom_point() +
    geom_errorbar(aes(xmin = `l-95% CI`, xmax = `u-95% CI`)) +
    geom_vline(xintercept = eff.x, color = "grey", linetype = "dashed") +
    labs(x = "Estimated effect strength", y = "Regression model") +
    guides(color = "none") +
    theme_bw()
```

Second, we can plot marginal effects.
It becomes very clear that the marginal effect inferred from the non-linear model is much closer to the actual distribution of the data.

```{r comparison-marginal}
ceff.linear <- conditional_effects(m.linear, effects = "x")[[1]] %>% 
  select(x, estimate__, lower__, upper__) %>% 
  mutate(model = "linear (x)")

ceff.nonlinear <- conditional_effects(m.nonlinear, effects = "x")[[1]] %>% 
  select(x, estimate__, lower__, upper__) %>% 
  mutate(model = "quadratic (x^2)")

rbind(ceff.linear, ceff.nonlinear) %>% 
  ggplot(aes(x = x)) +
    geom_line(aes(y = estimate__, color = model, group = model)) +
    geom_ribbon(aes(ymin = lower__, ymax = upper__, fill = model, group = model), alpha = 0.2) +
    labs(x = "x", y = "Estimation of y", color = "Model", fill = "Model") +
    theme_bw()
```

Finally, we can compare the predictive power of the two models.
The computed difference is strongly significant, supporting that the non-linear model estimates unseen data much better than the linear model.

```{r comparison-loo}
m.linear <- add_criterion(m.linear, "loo")
m.nonlinear <- add_criterion(m.nonlinear, "loo")

loo_compare(m.linear, m.nonlinear, criterion = "loo")
```

