---
title: "Causal DAGs"
author: "Julian Frattini"
date: '2024-10-15'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# cluster of libraries for easier R syntax
library(tidyverse)

# visualization and analysis of DAGs
library(ggdag)
```

## Visualizing Causal Assumptions

An essential step in statistical causal inference is to visualize causal assumptions.
An easy tool for this is a directed, acyclic graph (DAG), where nodes represents events (or variables) and edges represent (assumed) causal relationships.

### Specifying a DAG

Using the `ggdag::dagify` function, we can specify a minimal DAG that contains

1. a formula of the structure `response ~ treatment`,
2. an `exposure` that specifies the treatment variable of interest, and
3. an `outcome` that specifies the response variable of interest.

```{r dag1-specification}
dag1 <- dagify(
  a ~ b,
  exposure = "b",
  outcome = "a"
)
```

We can visualize the dag using the `ggdag::ggdag_status` function as such.

```{r dag1-visualization}
ggdag_status(dag1) +
  guides(fill = "none", color = "none") + # remove the "obvious" legends for fill and color
  theme_dag() # remove the axes and grid that ggplot defaults to
```

### Options

When specifying a DAG, we can make use of several additional options.
The following example demonstrates three of them: more than two variables and relationships, verbose labels, and specified structure.

```{r dag2}
dag2 <- dagify(
  # specify multiple relationships by providing more formulae (which may include more than 1 predictor)
  a ~ b + c + d,
  b ~ d + e,
  exposure = "b", outcome = "a",
  
  # specify labels by assigning one string to each variable used in the formulae
  labels = c(a = "performance", b = "programming language", c = "motivation", d = "environment", e = "education"),
  
  # specify coordinates by assigning one x and one y value to each variable in the formulae
  coords = list(x = c(a = 2, b = 1, c = 1, d = 1.5, e = 0), 
                y = c(a = 1, b = 1, c = 1.5, d = 0.5, e = 1))
)

ggdag_status(dag2, 
             text = FALSE, # disable the variable names (a, b, ...) within the nodes ...
             use_labels = "label") + # ... and enable the variable labels ("performance", ...)
  guides(fill = "none", color = "none") + 
  theme_dag()
```

Using these tools, even complex, hierarchical causal relationships can be specified and made explicit.
